---
title: "TimeSeries"
output: pdf_document
---
```{r}
#Libraries used
library(qpcR)
library(readxl)
library(forecast)
library(MASS)
library(readr)
library(ggplot2)
library(astsa)
library(tseries)
```

```{r}
#Importing code
ontario_excel <- read.csv("ontariocsv.csv")
death <- ontario_excel
death <- ts(death)
```


# creating initial time series
```{r}
plot(death, ylab = "Deaths", main = "Time Series of Monthly Traffic Fatalities in Ontario", col = "blue")
 # Exploratory Analysis
```
 
# Creating matrix of monthly accidents with years as rows and columns as months
 
 
```{r} 
year = matrix(NA, nrow = 15, ncol = 12)
rownames(year) <- paste(1960:1974)
colnames(year) <- paste(c('Jan', 'Feb', 'Mar', 'Apr', 'May', 'Jun', 'Jul', 'Aug', 'Sep', 'Oct', 'Nov', 'Dec'))
 
for (i in 0:14) {
  year[i+1,] <- death[((i*12)+1):(12*(i+1))]
}
 
year
```

# August and September tend to be the higher death months while January and February tend to be the lower death months.  This shows that there is a seasonal component to the time series.

Calculating sample mean for each year
```{r}
# Calculating sample mean for each year
mean.values = c(rep(0, 15))
for (i in 1:15) {
  mean.values[i] <- mean(year[i,])
}
mean.values
# There seems to be an increasing mean as each year increases.  This leads us to believe that there is a trend in the time series.
```

Calculating sample variance for each year
```{r}
var.values = c(rep(0, 15))
for (i in 1:15) {
  var.values[i] <- var(year[i,])
}
var.values
# There seems to be an increasing variance as each year increases.
```
Box-Cox Transformation
```{r}
bcTransform <- boxcox(death~ as.numeric(1:length(death)))
lambda.estimate <- bcTransform$x[which(bcTransform$y == max(bcTransform$y))]
 
 
lambda.estimate.values <- bcTransform$x[bcTransform$y > max(bcTransform$y) - 1/2 * qchisq(.95,1)] # Values above 95% CI line
lambda.lower.bound <- min(lambda.estimate.values)
lambda.upper.bound <- max(lambda.estimate.values)
 
lambda.estimate

lambda.lower.bound

lambda.upper.bound
```

Square Root Transformation
```{r}
death.sqrt <- sqrt(death)
var(death.sqrt) #variance of transformed time series
```

ACF and PACF of Square Root Transformed Data
```{r}
op <- par(mfrow = c(1,2))
ts.plot(death, main = "Original Data", ylab = expression(X[t]))
ts.plot(death.sqrt, main = "Square Root Transformed Data", ylab = expression(Y[t]))
 
par(op)
Diffencing
death.lag12 <- diff(death.sqrt, lag = 12)
 
death.lagged <- diff(death.lag12, lag = 1)
var(death.lagged)

var(death.lag12)  # differenced at lag 12

var(death.lagged) # then differenced at lag 1

# therefore not necessary to difference at lag 1 to remove trend because it increases variance
```

Dickey-Fuller Test
```{r}
adf.test(death.lag12)
```

Time Series of Differenced and transformed series and Times Sereis with just tranformation
```{r}
ts.plot(death.lag12, main = "De-seasonalized/Sqrt Ttransformed Time Series", ylab = expression(nabla^{12}~nabla~Y[t]))
abline(h=0, lty = 2)
 
ts.plot(death.sqrt, main = "Sqrt Transformed Time Series")
```

ACF/PACF of transformed and differenced time series
```{r}
# re-calculate the sample variance and examine the ACF and PACF
op = par(mfrow = c(1,2))
acf(death.lag12, lag.max = 40, main = "")
pacf(death.lag12, lag.max = 40, main = "")
title("", line = -1, outer=TRUE)
 
par(op)
```


AICc chart of possible models
```{r}
aiccs <- matrix(0,nrow = 5,ncol = 5)
dimnames(aiccs) <- list(p = 0:4, q = 0:4)
for (i in 0:4){
6
for (j in 0:4){
aiccs[i+1,j+1] <- AICc(arima(death.lag12,order = c(i,0,j),method = "ML"))
}
}
aiccs
```

Parameters of desired model type arima(1,0,1)
```{r}
fit <- Arima(death.lag12, order=c(1,0,1))
fit
```

Yule-Walker
```{r}
(fit <- ar(death.lag12, method = "yule-walker"))
```

Diagnostics
```{r}
i <- arima(death.lag12, order = c(1,0,1), method = "ML", xreg = 1:length(death.lag12))
 
residuals <- resid(i)
Checking for normality
plot(residuals)
 
hist(residuals, probability = T, main = "", xlab = "Residuals")
 
shapiro.test(residuals)

mean(residuals)

var(residuals)

qqnorm(residuals)
qqline(residuals, col = "red")
```

Checking for serial correlation
```{r}
Box.test(residuals, lag = 11, type = c("Ljung-Box"), fitdf = 2)
Box.test(residuals, lag = 11, type = c("Box-Pierce"), fitdf = 2)
```

Checking for heteroskedacity
```{r}
op <- par(mfrow = c(1,2))
acf(residuals, main = "", lag.max = 40)
pacf(residuals, main = "", lag.max = 40)
 
par(op)
#we have arma(1,0,1) model
```

Forecast Prediction 
```{r}
op = par(mfrow = c(1,1))
data_pred = predict(fit, n.ahead=12)
ts.plot(ontario, xlim = c(1,200), ylim=c(0,300), ylab="Deaths")
points(169:180, exp(data_pred$pred), col = "red", pch=20)
lines(169:180, exp(data_pred$pred+1.96*data_pred$se), lty=2)
lines(169:180, exp(data_pred$pred-1.96*data_pred$se), lty=2)

#legend
legend("topleft", legend = c("Predicted", "Observed"), col=c("red","black"), lty=1, cex = 0.7)
```

zoom in
```{r}
# zoom in at the predicted points
ts.plot(ontario, xlim=c(160, 185), ylim=c(10, 300), main = "Zoomed-in", ylab = "Deaths")
points(169:180, exp(data_pred$pred), col = "red", pch=20)
lines(169:180, exp(data_pred$pred+1.96*data_pred$se), lty=2)
lines(169:180, exp(data_pred$pred-1.96*data_pred$se), lty=2)

#legend
legend("bottomright", legend = c("Predicted", "Observed"), col=c("red","black"), lty=1, cex = 0.7)
```

fitted vs observed
```{r}
#observed and fitted plot
ts.plot(death, lwd=1, lty=1, main = "Fitted vs Observed")
lines(exp(fit$fitted), lwd=1, lty=1, col="red")

#legend
legend("bottomright", legend = "Fitted", col="red", lty=1, cex = 0.6)
```